---
title: "Data Science Specialization Capstone Project Exploratory Data Analysis"
author: "mspcvsp"
date: "Monday, November 10, 2014"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---  

```{r setupEnvironment, include=FALSE}
# http://stackoverflow.com/questions/13090838/r-markdown-avoiding-package-loading-messages
Sys.setenv(JAVA_HOME='C:\\Program Files\\Java\\jre1.8.0_20')

library(rJava)
library(RWeka)
library(R.utils)
library(tm)
library(stringi)
library(xtable)

if (!file.exists("./Data/final")) {
    unzip("./Data/Coursera-SwiftKey.zip", exdir="./Data")   
    
    # Profanity filter
    profanityBlackListURL <-
        paste0("http://www.frontgatemedia.com/new/wp-content/uploads/2014/03/",
               "Terms-to-Block.csv")
    
    download.file(url = profanityBlackListURL,
                  destfile="./Data/Terms-to-Block.csv")
}

source("./determineTextFileSize.R")
```

## Introduction

- The objective of the Data Science Capstone project is to develop an application that
predicts the next word a user will type into a mobile device
    - [Capstone Project Introduction](http://simplystatistics.org/2014/08/19/swiftkey-and-johns-hopkins-partner-for-data-science-specialization-capstone/)  
-  The purpose of this presentation is to discuss the following topics:
    - Summarize the statistics of the Capstone project text data sets
    - Describe unusual characteristics of this data  
- [An ISO 639-1 Language Code](http://www.w3schools.com/tags/ref_language_codes.asp)
    - Two-character string that refers to a language
    - Data Science Capstone project text data spans the following languages:  
        - English (EN)  
        - Finnish (FI)  
        - Germain (DE)  
        - Russian (RU)  
- [Data Science Capstone project Github repository](https://github.com/datasciencespm/DataScienceCapstone/tree/master)  

## Count the number of lines {.columns-2}
```{r, echo=FALSE, results='hide'}
baseTextFilePath <- "./Data/final"
baseOutputDataPath <- "./OutputData"

if (!file.exists(baseOutputDataPath)) {
    dir.create(baseOutputDataPath)
}

for (curLanguage in dir(baseTextFilePath)) {
    curOutputDatPath <- file.path(baseOutputDataPath, curLanguage)
    
    if (!file.exists(curOutputDatPath)) {
        dir.create(curOutputDatPath)
    }

    determineTextFileSize(file.path(baseTextFilePath, curLanguage),
                          curOutputDatPath)
}
```

- German text documents # of lines:  
```{r loadDENumberOfLines, echo=FALSE, results='asis'}
load("./OutputData//de_DE/de_DENumLines.RData")
num_lines = t(as.data.frame(num_lines))
colnames(num_lines) <- c("numberoflines")
print(xtable(num_lines),'html')
rm(num_lines)
```
  
- English text documents # of lines:  
```{r loadENNumberOfLines, echo=FALSE, results='asis'}
# http://blog.revolutionanalytics.com/2010/02/making-publicationready-tables-with-xtable.html
load("./OutputData//en_US/en_USNumLines.RData")
num_lines = t(as.data.frame(num_lines))
colnames(num_lines) <- c("numberoflines")
print(xtable(num_lines),'html')
rm(num_lines)
```
    
- Finnish text documents # of lines:
```{r loadFINumberOfLines, echo=FALSE, results='asis'}
# http://blog.revolutionanalytics.com/2010/02/making-publicationready-tables-with-xtable.html
load("./OutputData//fi_FI/fi_FINumLines.RData")
num_lines = t(as.data.frame(num_lines))
colnames(num_lines) <- c("numberoflines")
print(xtable(num_lines),'html')
rm(num_lines)
```
  
- Russian text documents # of lines:
```{r loadRUNumberOfLines, echo=FALSE, results='asis'}
# http://blog.revolutionanalytics.com/2010/02/making-publicationready-tables-with-xtable.html
load("./OutputData//ru_RU/ru_RUNumLines.RData")
num_lines = t(as.data.frame(num_lines))
colnames(num_lines) <- c("numberoflines")
print(xtable(num_lines),'html')
rm(num_lines)
```
